---
title: "Filtering raw inputs before RPCA"
date: "`r Sys.Date()`"
package: mia
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
        toc_depth: 2
        number_sections: true
vignette: >
    %\VignetteIndexEntry{Filtering raw inputs before RPCA}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
    cache = FALSE,
    message = FALSE,
    warning = FALSE
)
```

# Overview

`filterRPCAInput()` filters raw count assays before the robust centered
log-ratio transformation used by RPCA and Joint-RPCA. The function supports
both `SummarizedExperiment` and `MultiAssayExperiment` objects.

The three filtering arguments are:

* `min.sample.count`: retain samples whose total count is strictly greater
  than the threshold.
* `min.feature.count`: retain features whose total count is strictly greater
  than the threshold.
* `min.feature.frequency`: retain features whose percentage prevalence is
  strictly greater than the threshold.

Setting an argument to `NULL` disables that filter.

# Load *mia* and the example data

This example uses the internal `ibdmdb` dataset. The first two experiments
contain metagenomic (MGX) and metatranscriptomic (MTX) measurements.

```{r load-data}
library("mia")

data("ibdmdb", package = "mia")
mae <- ibdmdb

experiment_ids <- c(1L, 2L)
assay_types <- c("mgx", "mtx")
experiment_names <- names(
    MultiAssayExperiment::experiments(mae)
)[experiment_ids]
```

# Filter the raw count assays

Here, samples and features with zero total count are removed. Features must
also be present in more than 10% of the samples.

```{r filter-data}
filtered_mae <- filterRPCAInput(
    mae,
    experiments = experiment_ids,
    assay.types = assay_types,
    min.sample.count = 0,
    min.feature.count = 0,
    min.feature.frequency = 0
)
```

# Compare dimensions before and after filtering

```{r compare-dimensions}
before <- t(vapply(
    experiment_ids,
    function(i) {
        c(features = nrow(mae[[i]]), samples = ncol(mae[[i]]))
    },
    numeric(2L)
))

after <- t(vapply(
    experiment_ids,
    function(i) {
        c(
            features = nrow(filtered_mae[[i]]),
            samples = ncol(filtered_mae[[i]])
        )
    },
    numeric(2L)
))

data.frame(
    experiment = experiment_names,
    features_before = before[, "features"],
    features_after = after[, "features"],
    samples_before = before[, "samples"],
    samples_after = after[, "samples"],
    row.names = NULL
)
```
# Session info

```{r}
sessionInfo()
```




The returned object can then be used in the RPCA preprocessing workflow.
Filtering should be applied to raw counts, before creating an `rclr` assay.

For complete argument descriptions, see `?filterRPCAInput`.
