---
title: "Cross-implementation validation of Joint-RPCA"
subtitle: "Comparison of mia and Gemelli using paired IBDMDB MGX–MTX data"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show analysis code"
    embed-resources: true
    fig-responsive: true
    df-print: paged
engine: knitr
execute:
  echo: true
  warning: false
  message: false
  error: false
---

# Study objective

This notebook evaluates whether the Joint-RPCA implementation in the
development version of `mia` reproduces the numerical structure obtained with
Gemelli when both implementations are applied to the same paired MGX and MTX
data, use the same train–test split, and estimate the same number of latent
components.

Agreement is evaluated at five levels:

1. sample coordinates in the Joint-RPCA space;
2. feature loadings;
3. proportion of variation explained;
4. cross-validation error trajectories; and
5. pairwise sample distances.


# Interpretation guide

For the scatter plots, the solid diagonal line represents exact numerical
agreement. Points close to this line indicate that the two implementations
produce equivalent values. Correlation measures agreement in ordering and
relative structure, whereas deviations from the diagonal reveal differences
in scaling or absolute magnitude.

# Setup

```{r}
#| label: setup
#| include: false

project_root <- "D:/joint-rpca-local-validation"

knitr::opts_knit$set(root.dir = project_root)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(mia)
library(MultiAssayExperiment)
library(SummarizedExperiment)
library(ggplot2)
library(jsonlite)
library(knitr)
library(scales)

theme_set(
  theme_classic(base_size = 12) +
    theme(
      plot.title.position = "plot",
      plot.caption.position = "plot",
      plot.title = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(size = 10.5),
      plot.caption = element_text(size = 9, hjust = 0),
      axis.title = element_text(face = "plain"),
      strip.background = element_rect(fill = "grey95", colour = "grey70"),
      strip.text = element_text(face = "bold"),
      legend.position = "top"
    )
)

n_components <- 3L
max_iterations <- 10L
analysis_seed <- 20260630L

required_python_outputs <- file.path(
  project_root,
  "results",
  "gemelli",
  c(
    "python_versions.json",
    "sample_scores.csv",
    "feature_loadings.csv",
    "proportion_explained.csv",
    "distance_matrix.csv",
    "cv_error.csv"
  )
)
```

# Software provenance

```{r}
`%||%` <- function(x, fallback = NA_character_) {
  if (is.null(x) || length(x) == 0L || is.na(x)) fallback else x
}

mia_info <- packageDescription("mia")

mia_provenance <- data.frame(
  item = c(
    "mia version",
    "mia repository",
    "mia branch",
    "mia commit",
    "R version",
    "Platform"
  ),
  value = c(
    mia_info[["Version"]],
    paste0(
      mia_info[["RemoteUsername"]] %||% "microbiome",
      "/",
      mia_info[["RemoteRepo"]] %||% "mia"
    ),
    mia_info[["RemoteRef"]] %||% "devel",
    mia_info[["RemoteSha"]] %||% "not recorded",
    R.version.string,
    R.version$platform
  ),
  stringsAsFactors = FALSE
)

knitr::kable(mia_provenance)
```

```{r}
if (!file.exists(required_python_outputs[1])) {
  stop(
    "Python provenance is missing. Run: python python/export_python_versions.py",
    call. = FALSE
  )
}

python_versions <- jsonlite::read_json(
  required_python_outputs[1],
  simplifyVector = TRUE
)

python_provenance <- data.frame(
  software = names(python_versions),
  version = unname(unlist(python_versions)),
  stringsAsFactors = FALSE
)

knitr::kable(python_provenance)
```

# Data preparation

```{r}
data("ibdmdb", package = "mia")
mae <- ibdmdb

stopifnot(methods::is(mae, "MultiAssayExperiment"))
stopifnot(all(c("MGX", "MTX") %in% names(mae)))

mae <- MultiAssayExperiment::intersectColumns(mae)

mgx <- SummarizedExperiment::assay(mae[["MGX"]], "mgx")
mtx <- SummarizedExperiment::assay(mae[["MTX"]], "mtx")

stopifnot(
  identical(colnames(mgx), colnames(mtx)),
  all(is.finite(mgx)),
  all(is.finite(mtx)),
  all(mgx >= 0),
  all(mtx >= 0)
)
```

```{r}
rownames(mgx) <- paste0("MGX::", make.unique(rownames(mgx)))
rownames(mtx) <- paste0("MTX::", make.unique(rownames(mtx)))

rownames(mae[["MGX"]]) <- rownames(mgx)
rownames(mae[["MTX"]]) <- rownames(mtx)

stopifnot(anyDuplicated(c(rownames(mgx), rownames(mtx))) == 0L)

sample_ids <- colnames(mgx)

dataset_summary <- data.frame(
  modality = c("MGX", "MTX"),
  samples = c(ncol(mgx), ncol(mtx)),
  features = c(nrow(mgx), nrow(mtx)),
  zero_fraction = c(mean(mgx == 0), mean(mtx == 0))
)

dataset_summary$zero_fraction <- percent(
  dataset_summary$zero_fraction,
  accuracy = 0.1
)

names(dataset_summary) <- c(
  "Modality",
  "Samples",
  "Features",
  "Zero entries"
)

knitr::kable(
  dataset_summary,
  caption = "Characteristics of the two input assays",
  align = c("l", "r", "r", "r")
)
```

# Export of common inputs

```{r}
dir.create("data", recursive = TRUE, showWarnings = FALSE)
dir.create("results/mia", recursive = TRUE, showWarnings = FALSE)

write.csv(mgx, "data/mgx_raw.csv", quote = TRUE)
write.csv(mtx, "data/mtx_raw.csv", quote = TRUE)

set.seed(analysis_seed)
n_test <- max(1L, floor(0.20 * length(sample_ids)))
test_ids <- sort(sample(sample_ids, size = n_test, replace = FALSE))

split_df <- data.frame(
  sample_id = sample_ids,
  train_test = ifelse(sample_ids %in% test_ids, "test", "train"),
  stringsAsFactors = FALSE
)

write.csv(
  split_df,
  "data/train_test_split.csv",
  row.names = FALSE,
  quote = FALSE
)

stopifnot(
  file.exists("data/mgx_raw.csv"),
  file.exists("data/mtx_raw.csv"),
  file.exists("data/train_test_split.csv")
)

knitr::kable(as.data.frame(table(split_df$train_test)), col.names = c("Set", "Samples"))
```

# Joint-RPCA in R

```{r}
mae_rclr <- mae

mae_rclr[["MGX"]] <- transformAssay(
  mae_rclr[["MGX"]],
  assay.type = "mgx",
  method = "rclr",
  impute = FALSE,
  name = "rclr"
)

mae_rclr[["MTX"]] <- transformAssay(
  mae_rclr[["MTX"]],
  assay.type = "mtx",
  method = "rclr",
  impute = FALSE,
  name = "rclr"
)

stopifnot(
  "rclr" %in% SummarizedExperiment::assayNames(mae_rclr[["MGX"]]),
  "rclr" %in% SummarizedExperiment::assayNames(mae_rclr[["MTX"]])
)
```

```{r}
set.seed(analysis_seed)

mia_result <- mia::getJointRPCA(
  mae_rclr,
  experiments = c("MGX", "MTX"),
  assay.types = c("rclr", "rclr"),
  test.set = test_ids,
  ncomponents = n_components,
  max.iterations = max_iterations
)

mia_sample_scores <- as.matrix(mia_result)
mia_feature_loadings <- attr(mia_result, "rotation")
mia_proportion_explained <- as.numeric(attr(mia_result, "percentVar")) / 100
mia_distance <- as.matrix(dist(mia_sample_scores))

stopifnot(
  !anyNA(mia_sample_scores),
  nrow(mia_sample_scores) == length(sample_ids),
  ncol(mia_sample_scores) == n_components,
  ncol(mia_feature_loadings) == n_components
)

write.csv(mia_sample_scores, "results/mia/sample_scores.csv", quote = TRUE)
write.csv(mia_feature_loadings, "results/mia/feature_loadings.csv", quote = TRUE)
saveRDS(mia_result, "results/mia/mia_result.rds")
```

# Gemelli results

```{r}
missing_python_outputs <- required_python_outputs[!file.exists(required_python_outputs)]

if (length(missing_python_outputs) > 0L) {
  stop(
    "Gemelli outputs are missing: ",
    paste(missing_python_outputs, collapse = ", "),
    ". Run in WSL: python python/run_gemelli.py",
    call. = FALSE
  )
}

gemelli_sample_scores_raw <- as.matrix(read.csv(
  required_python_outputs[2],
  row.names = 1,
  check.names = FALSE
))

gemelli_feature_loadings_raw <- as.matrix(read.csv(
  required_python_outputs[3],
  row.names = 1,
  check.names = FALSE
))

gemelli_proportion_raw <- read.csv(
  required_python_outputs[4],
  row.names = 1,
  check.names = FALSE
)

gemelli_distance <- as.matrix(read.csv(
  required_python_outputs[5],
  row.names = 1,
  check.names = FALSE
))

gemelli_cv_error <- read.csv(
  required_python_outputs[6],
  row.names = 1,
  check.names = FALSE
)

mia_cv_error <- as.data.frame(attr(mia_result, "cv_error"))

stopifnot(
  setequal(rownames(mia_sample_scores), rownames(gemelli_sample_scores_raw)),
  setequal(rownames(mia_feature_loadings), rownames(gemelli_feature_loadings_raw)),
  nrow(mia_cv_error) > 0L,
  nrow(gemelli_cv_error) > 0L
)
```

# Alignment of components

```{r}
gemelli_sample_scores_raw <- gemelli_sample_scores_raw[
  rownames(mia_sample_scores),
  ,
  drop = FALSE
]

gemelli_feature_loadings_raw <- gemelli_feature_loadings_raw[
  rownames(mia_feature_loadings),
  ,
  drop = FALSE
]

component_correlations <- abs(cor(
  mia_sample_scores,
  gemelli_sample_scores_raw
))

permutations <- rbind(
  c(1L, 2L, 3L),
  c(1L, 3L, 2L),
  c(2L, 1L, 3L),
  c(2L, 3L, 1L),
  c(3L, 1L, 2L),
  c(3L, 2L, 1L)
)

permutation_scores <- apply(permutations, 1L, function(p) {
  sum(component_correlations[cbind(seq_len(n_components), p)])
})

best_permutation <- permutations[which.max(permutation_scores), ]

gemelli_sample_scores <- gemelli_sample_scores_raw[, best_permutation, drop = FALSE]
gemelli_feature_loadings <- gemelli_feature_loadings_raw[, best_permutation, drop = FALSE]
gemelli_proportion_explained <- as.numeric(gemelli_proportion_raw[, 1])[best_permutation]

for (j in seq_len(n_components)) {
  if (sum(mia_sample_scores[, j] * gemelli_sample_scores[, j]) < 0) {
    gemelli_sample_scores[, j] <- -gemelli_sample_scores[, j]
    gemelli_feature_loadings[, j] <- -gemelli_feature_loadings[, j]
  }
}

colnames(gemelli_sample_scores) <- colnames(mia_sample_scores)
colnames(gemelli_feature_loadings) <- colnames(mia_feature_loadings)
```


# Results

## Overall agreement summary

The following sections quantify agreement separately for each output of the
Joint-RPCA workflow. Sample scores are first aligned for component order, sign,
and component-specific scale. Feature loadings are aligned using the same
component order and sign convention.

```{r}
extract_cv_column <- function(x, candidates) {
  matched <- intersect(candidates, names(x))
  if (length(matched) == 0L) {
    stop(
      "Could not identify a CV column. Available columns: ",
      paste(names(x), collapse = ", "),
      call. = FALSE
    )
  }
  as.numeric(x[[matched[1]]])
}

spearman_label <- function(x, y) {
  test <- suppressWarnings(
    cor.test(x, y, method = "spearman", exact = FALSE)
  )

  rho <- unname(test$estimate)
  p_value <- test$p.value

  paste0(
    "Spearman rho=", format(round(rho, 4), nsmall = 1),
    ", p=", format(p_value, scientific = TRUE, digits = 2)
  )
}
```

## Sample scores

```{r}
# Component-specific scaling accounts for differing score normalization.
gemelli_scores_scaled <- gemelli_sample_scores
scale_factors <- numeric(n_components)

for (j in seq_len(n_components)) {
  scale_factors[j] <- sum(
    mia_sample_scores[, j] * gemelli_sample_scores[, j]
  ) / sum(gemelli_sample_scores[, j]^2)

  gemelli_scores_scaled[, j] <- gemelli_sample_scores[, j] * scale_factors[j]
}

pc_correlations <- diag(cor(mia_sample_scores, gemelli_sample_scores))
sample_score_table <- data.frame(
  Component = colnames(mia_sample_scores),
  `Pearson correlation` = pc_correlations,
  `mia SD` = apply(mia_sample_scores, 2, sd),
  `Gemelli SD before scaling` = apply(gemelli_sample_scores, 2, sd),
  `Applied scale factor` = scale_factors,
  check.names = FALSE
)

knitr::kable(
  sample_score_table,
  digits = 6,
  caption = paste(
    "Component-wise agreement of sample scores.",
    "Gemelli scores are shown before component-specific rescaling."
  ),
  align = c("l", "r", "r", "r", "r")
)
```

```{r}
sample_plot_data <- do.call(rbind, lapply(seq_len(n_components), function(j) {
  python_values <- gemelli_scores_scaled[, j]
  r_values <- mia_sample_scores[, j]

  data.frame(
    panel = spearman_label(python_values, r_values),
    component = paste0("PC", j),
    Python = python_values,
    R = r_values
  )
}))

ggplot(sample_plot_data, aes(x = Python, y = R)) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linewidth = 0.55,
    linetype = "dashed"
  ) +
  geom_point(size = 2.1, alpha = 0.75) +
  facet_wrap(
    vars(component, panel),
    scales = "free",
    nrow = 1,
    labeller = label_value
  ) +
  labs(
    title = "Agreement of Joint-RPCA sample scores",
    subtitle = paste(
      "Gemelli scores were aligned to mia for component order, sign,",
      "and component-specific scale"
    ),
    x = "Gemelli sample score",
    y = "mia sample score",
    caption = "Dashed line: exact agreement (y = x)."
  ) +
  theme(aspect.ratio = 1)
```

## Feature loadings

```{r}
feature_loading_correlations <- diag(cor(
  mia_feature_loadings,
  gemelli_feature_loadings
))

feature_loading_table <- data.frame(
  Component = paste0("PC", seq_len(n_components)),
  `Pearson correlation` = feature_loading_correlations,
  check.names = FALSE
)

knitr::kable(
  feature_loading_table,
  digits = 6,
  caption = "Component-wise agreement of feature loadings",
  align = c("l", "r")
)
```

```{r}
feature_plot_data <- do.call(rbind, lapply(seq_len(n_components), function(j) {
  python_values <- gemelli_feature_loadings[, j]
  r_values <- mia_feature_loadings[, j]

  data.frame(
    panel = spearman_label(python_values, r_values),
    component = paste0("PC", j),
    Python = python_values,
    R = r_values
  )
}))

ggplot(feature_plot_data, aes(x = Python, y = R)) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linewidth = 0.55,
    linetype = "dashed"
  ) +
  geom_point(size = 1.15, alpha = 0.45) +
  facet_wrap(
    vars(component, panel),
    scales = "free",
    nrow = 1,
    labeller = label_value
  ) +
  labs(
    title = "Agreement of Joint-RPCA feature loadings",
    subtitle = "Each point represents one MGX or MTX feature",
    x = "Gemelli feature loading",
    y = "mia feature loading",
    caption = "Dashed line: exact agreement (y = x)."
  ) +
  theme(aspect.ratio = 1)
```

## Proportion explained

```{r}
explained_variance_comparison <- data.frame(
  Component = paste0("PC", seq_len(n_components)),
  Gemelli = gemelli_proportion_explained,
  mia = mia_proportion_explained,
  check.names = FALSE
)

explained_variance_comparison$`Absolute difference` <- abs(
  explained_variance_comparison$Gemelli -
    explained_variance_comparison$mia
)

display_explained_variance <- explained_variance_comparison
display_explained_variance$Gemelli <- percent(
  display_explained_variance$Gemelli,
  accuracy = 0.01
)
display_explained_variance$mia <- percent(
  display_explained_variance$mia,
  accuracy = 0.01
)
display_explained_variance$`Absolute difference` <- percent(
  display_explained_variance$`Absolute difference`,
  accuracy = 0.0001
)

knitr::kable(
  display_explained_variance,
  caption = "Proportion of variation explained by each component",
  align = c("l", "r", "r", "r")
)
```

```{r}
explained_plot_data <- rbind(
  data.frame(
    Component = explained_variance_comparison$Component,
    Implementation = "Gemelli",
    Proportion = explained_variance_comparison$Gemelli
  ),
  data.frame(
    Component = explained_variance_comparison$Component,
    Implementation = "mia",
    Proportion = explained_variance_comparison$mia
  )
)

ggplot(
  explained_plot_data,
  aes(x = Component, y = Proportion, fill = Implementation)
) +
  geom_col(
    position = position_dodge(width = 0.72),
    width = 0.64
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Explained variation by Joint-RPCA component",
    subtitle = "Side-by-side comparison of Gemelli and mia",
    x = NULL,
    y = "Proportion explained"
  )
```

## Cross-validation results

```{r}
cv_n <- min(nrow(gemelli_cv_error), nrow(mia_cv_error))

gemelli_cv_mean <- extract_cv_column(
  gemelli_cv_error,
  c("mean_CV", "mean_cv", "mean")
)[seq_len(cv_n)]

gemelli_cv_sd <- extract_cv_column(
  gemelli_cv_error,
  c("std_CV", "std_cv", "sd_CV", "sd_cv", "std", "sd")
)[seq_len(cv_n)]

mia_cv_mean <- extract_cv_column(
  mia_cv_error,
  c("mean_CV", "mean_cv", "mean")
)[seq_len(cv_n)]

mia_cv_sd <- extract_cv_column(
  mia_cv_error,
  c("std_CV", "std_cv", "sd_CV", "sd_cv", "std", "sd")
)[seq_len(cv_n)]

cv_comparison <- data.frame(
  Iteration = seq_len(cv_n),
  `Python mean_CV` = gemelli_cv_mean,
  `Python std_CV` = gemelli_cv_sd,
  `R mean_CV` = mia_cv_mean,
  `R std_CV` = mia_cv_sd,
  check.names = FALSE
)

knitr::kable(
  cv_comparison,
  digits = 6,
  caption = "Cross-validation reconstruction-error summaries by iteration",
  align = c("r", "r", "r", "r", "r")
)
```

```{r}
cv_plot_data <- rbind(
  data.frame(
    metric = "mean_CV",
    Python = cv_comparison[["Python mean_CV"]],
    R = cv_comparison[["R mean_CV"]]
  ),
  data.frame(
    metric = "std_CV",
    Python = cv_comparison[["Python std_CV"]],
    R = cv_comparison[["R std_CV"]]
  )
)

cv_plot_data$panel <- ave(
  seq_len(nrow(cv_plot_data)),
  cv_plot_data$metric,
  FUN = function(index) {
    rep(
      spearman_label(
        cv_plot_data$Python[index],
        cv_plot_data$R[index]
      ),
      length(index)
    )
  }
)

ggplot(cv_plot_data, aes(x = Python, y = R)) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linewidth = 0.55,
    linetype = "dashed"
  ) +
  geom_point(size = 2.3, alpha = 0.8) +
  facet_wrap(
    vars(metric, panel),
    scales = "free",
    nrow = 1,
    labeller = label_value
  ) +
  labs(
    title = "Agreement of cross-validation error summaries",
    subtitle = "Each point represents one optimization iteration",
    x = "Gemelli cross-validation metric",
    y = "mia cross-validation metric",
    caption = "Dashed line: exact agreement (y = x)."
  ) +
  theme(aspect.ratio = 1)
```

## Distance structure

```{r}
gemelli_distance <- gemelli_distance[
  rownames(mia_distance),
  colnames(mia_distance),
  drop = FALSE
]

upper_index <- upper.tri(mia_distance, diag = FALSE)

distance_correlation <- cor(
  mia_distance[upper_index],
  gemelli_distance[upper_index]
)

distance_comparison <- data.frame(
  Metric = "Pearson correlation",
  Value = distance_correlation
)

knitr::kable(
  distance_comparison,
  digits = 6,
  caption = "Agreement between pairwise sample-distance matrices",
  align = c("l", "r")
)
```

```{r}
distance_plot_data <- data.frame(
  Gemelli = gemelli_distance[upper_index],
  mia = mia_distance[upper_index]
)

ggplot(distance_plot_data, aes(x = Gemelli, y = mia)) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linewidth = 0.55,
    linetype = "dashed"
  ) +
  geom_point(size = 1.5, alpha = 0.55) +
  coord_equal() +
  labs(
    title = "Agreement of pairwise sample distances",
    subtitle = paste0(
      "Pearson correlation = ",
      format(distance_correlation, digits = 6)
    ),
    x = "Gemelli pairwise distance",
    y = "mia pairwise distance",
    caption = paste(
      "Each point is one unique sample pair;",
      "the dashed line denotes exact agreement."
    )
  )
```

# Main findings

```{r}
overall_summary <- data.frame(
  Quantity = c(
    "Minimum sample-score correlation",
    "Minimum feature-loading correlation",
    "Maximum explained-variation difference",
    "Distance-matrix correlation"
  ),
  Value = c(
    min(pc_correlations),
    min(feature_loading_correlations),
    max(explained_variance_comparison$`Absolute difference`),
    distance_correlation
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  overall_summary,
  digits = 8,
  caption = "Numerical summary of cross-implementation agreement",
  align = c("l", "r")
)
```




# Session information

```{r}
sessionInfo()
```
