---
title: "mia: Microbiome analysis tools"
date: "`r Sys.Date()`"
package: mia
output:
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: yes
        toc_depth: 2
        number_sections: true
vignette: >
    %\VignetteIndexEntry{mia}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, echo=FALSE}
knitr::opts_chunk$set(cache=FALSE,
                      fig.width=9,
                      message=FALSE,
                      warning=FALSE)
```

# Load *mia*

```{r load-packages, message=FALSE, warning=FALSE}
library("mia")
```

# Loading a `MicrobiomeExperiment` object

```{r}
data(GlobalPatterns)
me <- as(GlobalPatterns,"MicrobiomeExperiment")
me
```

# Functions for working microbiome data

## Merging and agglomeration based on taxonomic information.

In addition to the `aggValue` functions provide by `TreeSummarizedExperiment`
`agglomerateByRank` is available if, tree construction proves difficult.

`agglomerateByRank` constructs a `factor` to guide merging from the available
taxonomic information. For more information on merging have a look at the man
page via `?mergeRows`.

```{r}
# print the available taxonomic ranks
colnames(rowData(me))
taxonomyRanks(me)

# agglomerate at the Family taxonomic rank
x1 <- agglomerateByRank(me, rank = "Family")
## How many taxa before/after agglomeration?
nrow(me)
nrow(x1)

# with agglomeration of the tree
x2 <- agglomerateByRank(me, rank = "Family",
                        agglomerateTree = TRUE)
nrow(x2) # same number of rows, but
rowTree(x1) # ... different
rowTree(x2) # ... tree

## Look at enterotype dataset...
data(enterotype)
## print the available taxonomic ranks. Shows only 1 rank available
## not useful for agglomerateByRank
taxonomyRanks(enterotype)
```

To keep data tidy, the agglomerated data can be stored as an alternative 
experiment in the object of origin. With this synchronized sample subseting
becomes very easy.

```{r}
altExp(me, "family") <- x2
```

Keep in mind, that rows with `NA` or similar value (via the 
`empty.fields` argument) will be removed, if you set `na.rm = FALSE`. Depending
on this different number of rows will be returned.

```{r}
x1 <- agglomerateByRank(me, rank = "Species", na.rm = TRUE)
altExp(me,"species") <- agglomerateByRank(me, rank = "Species", na.rm = FALSE)
dim(x1)
dim(altExp(me,"species"))
```

## Constructing a tree from taxonomic data

```{r}
taxa <- rowData(altExp(me,"species"))[,taxonomyRanks(me)]
taxa_res <- resolveLoop(as.data.frame(taxa))
taxa_tree <- toTree(data = taxa_res)
taxa_tree$tip.label <- getTaxonomyLabels(altExp(me,"species"))
rowNodeLab <- getTaxonomyLabels(altExp(me,"species"), make_unique = FALSE)
altExp(me,"species") <- changeTree(altExp(me,"species"),
                                   rowTree = taxa_tree,
                                   rowNodeLab = rowNodeLab)
```

# Utility functions

To make migration and adoption as easy as possible several utility functions
are available.

## General wrappers

`getAbundanceFeature` and `getAbundanceSample` are wrappers on row-wise or 
column-wise assay data subsetting.

```{r}
abund <- getAbundanceSample(me, "CC1", abund_values = "counts")
all(abund == assay(me, "counts")[,"CC1"])
abund <- getAbundanceFeature(me, "522457", abund_values = "counts")
all(abund == assay(me, "counts")["522457",])
```

## Selecting most interesing features 

`getTopTaxa` returns a vector of the most `top` abundant feature IDs.

```{r}
top_taxa <- getTopTaxa(me,
                       method="mean",
                       top=5,
                       abund_values="counts")
top_taxa
```

## Generating tidy data

To generate tidy data as used and required in most of the tidyverse, `meltAssay`

```{r}
molten_data <- meltAssay(me,
                         add_row_data = TRUE,
                         add_col_data = TRUE,
                         abund_values = "counts")
molten_data
```

